Higher-Order unification for free!

### 2024-05-02 paper/main.tex
update constraint code snippet
minor
proofreading

\section{Introduction}
\label{sec:intro}
We aim to implement a
form of proof search known as type-class resolution~\cite{wadler89,sozeau08}
for the Coq's system~\cite{Coq-refman}.
Type-class solvers are unification based proof search procedures
reminiscent of Prolog, which back-chain lemmas taken
from a database of ``type-class instances''. Given this
analogy with Logic Programming we want to leverage the
Elpi meta-programming language~\cite{tassi:hal-01637063},
a dialect of $\lambda$Prolog, already used to extend
Coq in various ways~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.

@@

\noindent
The proof found by the solver back-chains on rule 3 (the only rule
about the $\forall$ quantifier), and then solves the premises with
rules 1 and 2 respectively.
Note that rule 3 features a second-order parameter \coqIn{P} that represents
a function of type \coqIn{A ~$\to$~ Prop} (a 1-argument predicate over \coqIn{A}).
The solver has to infer a value for \coqIn{P} by unifying the conclusion
of rule 3 with the goal, and in particular, it has to solve the unification
problem \coqIn{P x = nfact x 3}. This higher-order problem falls in the so-called 
pattern fragment and admits a unique
solution $\customlabel{solution:intro}{\rho}$ that assigns
the term \elpiIn{~$\lambda$~x.nfact x 3} to \coqIn{P}.
In the rest of this paper, we refer to the pattern fragment as
\llambda~\cite{miller92jsc}.

@@

\noindent
Unfortunately this encoding of rule \ref{clause:r3} does not work
since it uses the predicate \coqIn{P} as a first order term: for the meta
language its type is \elpiIn{tm}. If we try to back-chain the rule
\ref{clause:r3} on the encoding of the goal \ref{goal:g} given below

@@

\noindent
The root cause of the problems we outlined in this example
is a subtle mismatch between the equational theories of the meta language
and the object language, which in turn makes the 
unification procedures of the meta language weak.
The equational theory of the meta-language Elpi encompasses
$\eta\beta$-equivalence and its unification procedure can solve higher-order
problems in the pattern fragment. Although 
the equational theory of Coq is much richer, for efficiency and predictability
reasons, automatic proof search procedures typically employ a unification
procedure that only captures a $\eta\beta$-equivalence and only operates
in \llambda. The similarity is striking, but one needs to exercise some caution in order to
simulate a higher-order logic program in Coq using the unification of Elpi.

@@

To reason about unification, we provide a description of the
\Fo{} and \Ho languages where unification variables
are first-class terms, i.e. they have a concrete syntax as
shown in \cref{code:common-terms}.
Unification variables
in \Fo{} (\elpiIn{fuva} term constructor) have no explicit scope:
the arguments of a higher-order variable are given via the \elpiIn{fapp}
constructor. For example the term <<\coqIn{P x}>> is represented as
<<\elpiIn{fapp [fuva N, x]}>>, where \elpiIn{N} is the memory address
of \elpiIn{P} and \elpiIn{x} is a bound variable.\\
In \Ho, the representation of <<\coqIn{P x}>> is instead <<\elpiIn{uva N [x]}>>,
since unification variables are higher-order and come equipped with an
explicit scope.
%
% We keep these languages
%minimal, for example, we omit the \elpiIn{all} quantifier of CIC we used
%in the example in Section~\ref{sec:intro} together with the type notation of
%terms carried by the \elpiIn{lam} constructor.
%
{
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{-13pt}
\begin{figure}[b] %[H]
  \begin{tabular}{ll}
  \begin{minipage}{0.21\textwidth}
   \inputrawelpicode{code/fo_tm}
  \end{minipage}
  &
  \begin{minipage}{0.24\textwidth}
   \inputrawelpicode{code/ho_tm}
  \end{minipage}
  \end{tabular}\vspace{4pt}
  \caption{The \Fo{} and \Ho languages}\vspace{0.3em}
  \label{code:common-terms}
  \Description[code:common-terms]{code:common-terms}
\end{figure}
}
%\todo{use faddr for Fo}
%
%faddr and addr are memory addresses, the details are given in bla,
%here they are just unique identifiers for unif variables and their types
%do now allow to mistake one for the other.
%\todo{cleanup}

@@

\subsubsection{Term equality: \Eo and \Ee}
For both languages, we extend the equational theory
over ground terms to the full language by adding reflexivity for
unification variables, i.e. a unification variable is equal to itself.

@@

\noindent
Note that the symbol \elpiIn{|} separates the head of a list from the tail;
that \elpiIn{forall2} applies a binary predicate to two lists (of same
length); and that \elpiIn{=>} is $\lambda$Prolog syntax for
implication an is used to augment the
current program with an extra rule. Both procedures use implication to
assume that the fresh nominal constants introduces by \elpiIn{pi} are
equal to themselves.

@@

\section{Handling of \maybeeta}\label{sec:eta}
A term 
$\lambda x. t \appsep x$ is said to be the $\eta$-expansion of $t$ if
$x$ does not occur free in $t$, and conversely, $t$ is the $\eta$-contraction of
$\lambda x. t \appsep x$. The equational theory of \Fo{} identifies these terms,
but the current compilation scheme does not,
as shown by the following example:
while $\lambda x.X \appsep x \Uo{} f$ does admit the solution
$\rho = \{~ X \mapsto f ~\}$, the corresponding problem in
\hoUnifPb does not.
%
\printAlll
  {{{\lambda x. X \appsep x, f}}}
  {{{\lambda x. A_x, f}}}
  {{{X,A,1}}}
  {{}}

@@

\begin{elpicode}
link-eta L R :- var L, maybe-eta R, !,
  declare_constraint (link-eta L R) [L,R].
link-eta L R :- not (var L), !, eta-progress-lhs L R.
link-eta L R :- not (maybe-eta R), !, eta-progress-rhs L R.
\end{elpicode}

>>>>

\section{Introduction}
\label{sec:intro}
We aim to implement a
form of proof search known as type-class resolution~\cite{wadler89,sozeau08}
for the Coq's system~\cite{Coq-refman}.
Type-class solvers are unification based proof search procedures
reminiscent of Prolog, which back-chain lemmas taken
from a database of ``type-class instances''. Given this
analogy with Logic Programming we want to leverage the
Elpi meta programming language~\cite{tassi:hal-01637063},
a dialect of $\lambda$Prolog, already used to extend
Coq in various ways~\cite{tassi:hal-01637063,tassi:hal-01897468,gregoire:hal-03800154,newtc}.

@@

\noindent
The proof found by the solver back-chains on rule 3 (the only rule
about the $\forall$ quantifier), and then solves the premises with
rules 1 and 2 respectively.
Note that rule 3 features a second-order parameter \coqIn{P} that represents
a function of type \coqIn{A ~$\to$~ Prop} (a 1-argument predicate over \coqIn{A}).
The solver infers a value for \coqIn{P} by unifying the conclusion
of rule 3 with the goal, and in particular, it has to solve the unification
problem \coqIn{P x = nfact x 3}. This higher-order problem falls in the so-called 
pattern fragment and admits a unique solution:
\begin{elpicode}
P = ~$\lambda$~x.nfact x 3                                       ~\customlabel{solution:intro}{(\rho)}~
\end{elpicode}
In the rest of this paper, we refer to the pattern fragment as
\llambda~\cite{miller92jsc}.

@@

\noindent
Unfortunately this encoding of rule \ref{clause:r3} does not work
since it uses the predicate \coqIn{P} as a first order term: for the meta
language its type is \elpiIn{tm}. If we try to back-chain the rule
\ref{clause:r3} on the encoding of the goal \ref{goal:g} given here

@@

\noindent
The root cause of the problems we outlined in this example
is a subtle mismatch between the equational theories of the meta language
and the object language, which in turn makes the 
unification procedures of the meta language weak.
The equational theory of the meta-language Elpi encompasses
$\eta\beta$-equivalence and its unification procedure can solve higher-order
problems in the pattern fragment. Although 
the equational theory of Coq is much richer, for efficiency and predictability
reasons automatic proof search procedures typically employ a unification
procedure that only captures a $\eta\beta$-equivalence and only operates
in \llambda. The similarity is striking, but one needs to exercise some caution in order to
simulate a higher-order logic program in Coq using the unification of Elpi.

@@

In order to reason about unification we provide a description of the
\Fo{} and \Ho languages where unification variables
are first-class terms, i.e. they have a concrete syntax as
shown in \cref{code:common-terms}.
Unification variables
in \Fo{} (\elpiIn{fuva} term constructor) have no explicit scope:
the ``arguments'' of a higher-order variable are given via the \elpiIn{fapp}
constructor. For example the term <<\coqIn{P x}>> is represented as
<<\elpiIn{fapp [fuva N, x]}>>, where \elpiIn{N} is the memory address
of \elpiIn{P} and \elpiIn{x} is a bound variable.
In \Ho{} the representation of <<\coqIn{P x}>> is instead <<\elpiIn{uva N [x]}>>,
since unification variables are higher-order and come equipped with an
explicit scope.
%
% We keep these languages
%minimal, for example, we omit the \elpiIn{all} quantifier of CIC we used
%in the example in Section~\ref{sec:intro} together with the type notation of
%terms carried by the \elpiIn{lam} constructor.
%
{
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{-13pt}
\begin{figure}[b] %[H]
  \begin{tabular}{ll}
  \begin{minipage}{0.21\textwidth}
   \inputrawelpicode{code/fo_tm}
  \end{minipage}
  &
  \begin{minipage}{0.24\textwidth}
   \inputrawelpicode{code/ho_tm}
  \end{minipage}
  \end{tabular}\vspace{4pt}
  \caption{The \Fo{} and \Ho languages}\vspace{0.3em}
  \label{code:common-terms}
  \Description[code:common-terms]{code:common-terms}
\end{figure}
}
%\todo{use faddr for Fo}
%
%faddr and addr are memory addresses, the details are given in bla,
%here they are just unique identifiers for unif variables and their types
%do now allow to mistake one for the other.
%\todo{cleanup}

@@

\subsubsection{Term equality: \Eo and \Ee}
For both languages, we extend the equational theory
over ground terms to the full language by adding reflexivity for
unification variables, i.e. a unification variable is equal to itself
but different to any other term.

@@

\noindent
Note that the symbol \elpiIn{|} separates the head of a list from the tail;
that \elpiIn{forall2} applies a binary predicate to two lists (of equal
length); and that \elpiIn{=>} is $\lambda$Prolog syntax for
implication and is used to augment the
current program with an extra rule. Both procedures use implication to
assume that the fresh nominal constants introduces by \elpiIn{pi} are
equal to themselves.

@@

\section{Handling of \maybeeta}\label{sec:eta}
A term 
$\lambda x. t \appsep x$ is said to be the $\eta$-expansion of $t$ if
$x$ does not occur free in $t$, and conversely we say that $t$ is the $\eta$-contraction of
$\lambda x. t \appsep x$. The equational theory of \Fo{} identifies these terms,
but the current compilation scheme does not,
as shown by the following example:
while $\lambda x.X \appsep x \Uo{} f$ does admit the solution
$\rho = \{~ X \mapsto f ~\}$, the corresponding problem in
\hoUnifPb does not.
%
\printAlll
  {{{\lambda x. X \appsep x, f}}}
  {{{\lambda x. A_x, f}}}
  {{{X,A,1}}}
  {{}}

@@

\begin{elpicode}
link-eta L R :- not (var L), !, eta-progress-lhs L R.
link-eta L R :- not (maybe-eta R), !, eta-progress-rhs L R.
link-eta L R :- declare_constraint (link-eta L R) [L,R].
\end{elpicode}
